Stage-Splitting a Modal Language

### 2014-10-20 paper/example.tex
first pass on example.tex

\item
I don't like fuzzy words like hygenic that dont mean anything.
  What is the point of using them?  A paper ought to be written for
  the reader not as a note to self.
\end{itemize}

@@

Continuing on our example, consider an application where we perform
many order statistics queries using the same collection~\texttt{l}, but
potentially with different ranks~\cite{k}.  To take advantage of this,
we can {\em stage} quickselect so that the parts of the computation
that depend on the collection~\texttt{l} is performed once, and the
desired result is computed by using~\texttt{k} and the results from
this prior stage.

Prior work proposed staging for exactly these kinds of applications
and proposed techniques for writing staged
programs~\cite{staging}. Using techniques essentially the same as
prior work, we can stage our quickselect code by using staged typing
and inserting three keywords~\texttt{next, prev, hold} into the
appropriate places in the code as shown in \figref{qs-staged}. To
write the staged version, the programmer first observes that the
rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
rather than $\rmint$, indicating that it will be available in the
second stage.  All types that are not of circle type are available in
the first stage.  Naturally the output type, representing the $k$th
largest element of the list, is also $\fut\rmint$, since it is not
computed until the second stage. Having determined the type signature,
the programmer can then insert $\next$ and $\prev$ constructs into the
code. The construct $\next$ occurs in a stage \bbone\ context and
indicates that the contents of its block are stage \bbtwo, whereas
$\prev$ occurs in a stage \bbtwo\ context and indicates that the
contents of its block is stage \bbone.  (We adopt the convention that
the top-level context is stage \bbone.)  The output type of a $\next$
block is the $\fut$'d version of the type of its stage
\bbtwo\ contents.  For example, \verb|next{0}| from above has the type
$\fut\rmint$.  Correspondingly, $\prev$ requires that its stage
\bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
For example, \verb|prev{k}| from above has type $\rmint$ at stage
\bbtwo, since $k$ is bound to a $\fut\rmint$ at stage
\bbone. Informally speaking, the
$\next$ construct enables operating on second stage computations in
the first stage essentially by treating them as boxes that can be
moved around, stored in data structures, and passed to functions, and
the $\prev$ construct enables unwrapping the contents of such a box to
use in the second stage.

For convenience, the code also uses $\pause$ blocks.  This construct
has same stage signature as $\prev$, but instead of ``unwrapping"
$\fut$ types it simply promotes integers from stage~\bbone\ to
stage~\bbtwo.
\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
We instead provide it as a primitive to shorten examples.  
Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
This is related to the notion of {\em mobility} in \cite{murphy05} and
{\em stability} in \cite{krishnaswami13}.}

We note that given the type signature of staged quickselect, there are
several ways to insert the staging annotations.  The version we
presented here maximizes the work done at stage~\bbone.

@@

The benefit of this ``optimization'' can be dramatic.  For example, in
principle at least, the first stage function \texttt{quickselect1}
could produce a balanced binary search tree holding the elements
of~\texttt{l} and the second stage function \texttt{quickselect2}
could use this balanced binary search tree and the given rank value to
perform a logarithmic-time search in the tree to return the desired
result.  Assuming that the list~\texttt{l} contains $n$ elements, this
optimization would change the asymptotic complexity from expected
(randomized) $\Theta(n \cdot m)$ to $\Theta(n\log{n} \cdot m
\cdot\log{n})$, which for any $m \approx n$ reduces the complexity
from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
improvement.  Note that the desired optimized code shows above is
intellectually more sophisticated than the code that we have started
with: the optimized code is able to create a data structure, a
balanced binary tree augmented with indexing information, and use a
binary search technique over this tree to compute the result
asymptotically more efficiently.  In fact, based our teaching
experience, we can imagine this kind of problem to be a moderately
difficult exam question in an undergraduate algorithms class, as it
not only requires understanding of data structures such as binary
search trees but also requires modifying them to augment with indexing
information to support rank-based search.

>>>>

\item
I don't like fuzzy words like hygenic that dont mean anything.
  What is the point of using them?  A paper ought to be written for
  the reader not as a note to self.

\item the splitting example should shed more light to the techniques
  and try to give more insight.
\end{itemize}

@@

Continuing on our example, consider an application where we perform
many order statistics queries using the same collection~\texttt{l}, but
 with $m$ different ranks~\texttt{k}:
%
\begin{lstlisting}
quickselect l 1
quickselect l 2
 @$\ldots$@ 
quickselect l @$m$@.
\end{lstlisting}
%
Since \texttt{quickselect1} requires only on a first stage
argument~\texttt{l}, we can run it once to produce a boundary data
structure that can be used for all subsequent queries on~\texttt{l},
e.g., the code consisting of $m$ calls To take advantage of this, we
can {\em stage} quickselect so that the parts of the computation that
depend on the collection~\texttt{l} is performed once, and the desired
result is computed by using~\texttt{k} and the results from this prior
stage.

Prior work proposed staging for exactly these kinds of applications
and proposed techniques for writing staged
programs~\cite{staging}. Using techniques essentially the same as
prior work, we can stage our quickselect code by using staged typing
and inserting three keywords~\texttt{next, prev, hold} into the
appropriate places in the code as shown in \figref{qs-staged}. To
write the staged version, the programmer first observes that the
rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
rather than $\rmint$, indicating that it will be available in the
second stage.  All types that are not of circle type are available in
the first stage.  Naturally the output type, representing the $k$th
largest element of the list, is also $\fut\rmint$, since it is not
computed until the second stage.

Having determined the type signature,
the programmer can then insert $\next$ and $\prev$ constructs into the
code. The construct $\next$ occurs in a stage \bbone\ context and
indicates that the contents of its block are stage \bbtwo, whereas
$\prev$ occurs in a stage \bbtwo\ context and indicates that the
contents of its block is stage \bbone.  (We adopt the convention that
the top-level context is stage \bbone.)  The output type of a $\next$
block is the $\fut$'d version of the type of its stage
\bbtwo\ contents.  For example, \verb|next{0}| from above has the type
$\fut\rmint$.  Correspondingly, $\prev$ requires that its stage
\bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
For example, \verb|prev{k}| from above has type $\rmint$ at stage
\bbtwo, since $k$ is bound to a $\fut\rmint$ at stage
\bbone. Informally speaking, the
$\next$ construct enables operating on second stage computations in
the first stage essentially by treating them as boxes that can be
moved around, stored in data structures, and passed to functions, and
the $\prev$ construct enables unwrapping the contents of such a box to
use in the second stage.

For convenience, the code also uses $\pause$ blocks.  This construct
has same stage signature as $\prev$, but instead of ``unwrapping"
$\fut$ types it simply promotes integers from stage~\bbone\ to
stage~\bbtwo.
\footnote{It will turn out that $\pause$ is implementable---though it
  takes some effort---from our other language features.  We instead
  provide it as a primitive to shorten examples.  Furthermore, it
  would be wise to extend $\pause$ to other base types, if we had
  them, and to products and sums thereof.  This is related to the
  notion of {\em mobility} in \cite{murphy05} and {\em stability} in
  \cite{krishnaswami13}.} We note that given the type signature of
staged quickselect, there are several ways to insert the staging
annotations.  The version we presented here maximizes the work done at
stage~\bbone.

@@

The benefit of this ``optimization'' can be dramatic.  For example, in
principle at least, the first stage function \texttt{quickselect1}
could produce a balanced binary search tree holding the elements
of~\texttt{l} and the second stage function \texttt{quickselect2}
could use this balanced binary search tree and the given rank value to
perform a logarithmic-time search in the tree to return the desired
result.  Assuming that the list~\texttt{l} contains $n$ elements, this
optimization would change the asymptotic complexity from expected
(randomized) $\Theta(n \cdot m)$ to $\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the complexity
from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
improvement.  Note that the desired optimized code shows above is
intellectually more sophisticated than the code that we have started
with: the optimized code is able to create a data structure, a
balanced binary tree augmented with indexing information, and use a
binary search technique over this tree to compute the result
asymptotically more efficiently.  In fact, based our teaching
experience, we can imagine this kind of problem to be a moderately
difficult exam question in an undergraduate algorithms class, as it
not only requires understanding of data structures such as binary
search trees but also requires modifying them to augment with indexing
information to support rank-based search.
